Annotated example of index standardization with VAST

### 2021-12-14 doc/appendix-pcod-VAST.Rmd
Develop VAST comparison

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).  Several options are built in, for commonly-used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.

@@

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # this is for conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = FALSE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- as.numeric(1)

fit <- fit_model(
  settings = settings,
  Lat_i = pcod[, "lat"],
  Lon_i = pcod[, "lon"],
  t_i = pcod[, "year"],
  b_i = pcod[, "density"],
  a_i = pcod[, "effort"],
  input_grid = input_grid,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

We can look at parameter estimates and their standard errors.

@@

While making custom plots of individual elements requires a fair bit of work to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want. While here we stick with the default set of plots, where `plot_set = 3`, one can specify different standard plots to make by changing the setting of this argument (see `?FishStatsUtils::plot_maps`).

```{r save-plots, echo = TRUE, warning = FALSE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

@@

```{r plot-knots, echo = TRUE}
knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE}
knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
```

@@

>>>>

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function `fit_model()` and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST).  Several options are built in for commonly used regions, but in this example we'll need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in make_setting and `"input_grid" = input_grid` in `fit_model()`.

@@

settings <- make_settings(
  n_x = 205, # number of vertices in the SPDE mesh
  Region = "User",
  purpose = "index2", # index of abundance with Gamma for positive catches
  fine_scale = TRUE, # use bilinear interpolation from the INLA 'A' matrix
  zone = 9,
  FieldConfig = FieldConfig,
  RhoConfig = RhoConfig,
  ObsModel = c(2, 0), # conventional logit-linked delta-Gamma; c(10, 2) for Tweedie
  bias.correct = TRUE,
  use_anisotropy = FALSE,
  max_cells = Inf, # use all grid cells from the extrapolation grid
  knot_method = "grid" # or "samples"
)
```

@@

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
f <- here("doc", "appendix-VAST", "vast-cache.rds")
if (!file.exists(f)) {
  fit <- fit_model(
    settings = settings,
    Lat_i = pcod[, "lat"],
    Lon_i = pcod[, "lon"],
    t_i = pcod[, "year"],
    b_i = pcod[, "density"],
    a_i = pcod[, "effort"],
    input_grid = input_grid,
    working_dir = paste0(here("doc", "appendix-VAST"), "/")
  )
  saveRDS(fit, file = f)
} else {
  fit <- readRDS(f)
}
```

We can look at parameter estimates and their standard errors. First we see estimates from the binomial component and second we see estimates from the positive Gamma component.

@@

sdmTMB does not have built-in delta models; however, we can fit the two components separately and combine the predictions. Here, we will fit the binomial presence-absence model and then a model for the positive catches only with a Gamma observation distribution and a log link. We will use the VAST-created mesh so the results are comparable.

```{r sdmTMB-example, echo = TRUE, cache=TRUE}
mesh <- make_mesh(pcod, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)
m1 <- sdmTMB(present ~ 0 + as.factor(year), 
  data = pcod, mesh = mesh, family = binomial(link = "logit"),
  time = "year", spatiotemporal = "iid", spatial = "on")
m1

dat2 <- subset(pcod, density > 0)
mesh2 <- make_mesh(dat2, xy_cols = c("X", "Y"),
  mesh = fit$spatial_list$MeshList$isotropic_mesh)
m2 <- sdmTMB(density ~ 0 + as.factor(year), 
  data = dat2, mesh = mesh2, family = Gamma(link = "log"),
  time = "year", spatiotemporal = "iid", spatial = "on")
m2
```

```{r compare-vast-sdmTMB1, echo = FALSE}
plot_betas <- function(vast_model, sdmTMB_model, vast_par = "beta1_ft") {
  s <- vast_model$parameter_estimates$SD
  vast_est1 <- as.list(s, "Estimate", report = FALSE)
  vast_est2 <- as.list(s, "Estimate", report = TRUE)
  vast_sd1 <- as.list(s, "Std. Error", report = FALSE)
  vast_sd2 <- as.list(s, "Std. Error", report = TRUE)
  sdmTMB_est <- as.list(sdmTMB_model$sd_report, "Estimate", report = FALSE)
  sdmTMB_sd <- as.list(sdmTMB_model$sd_report, "Std. Error", report = FALSE)
  b_year_vast <- vast_est1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  b_year_vast_se <- vast_sd1[[vast_par]][!is.na(vast_sd1[[vast_par]])]
  years <- sort(unique(pcod$year))
  lwr_vast <- b_year_vast - 2 * b_year_vast_se
  upr_vast <- b_year_vast + 2 * b_year_vast_se
  plot(years, b_year_vast, ylim = range(c(lwr_vast, upr_vast)))
  segments(years, lwr_vast, years, upr_vast)
  years <- years + 0.05
  points(years, sdmTMB_est$b_j)
  segments(years, sdmTMB_est$b_j - 2 * sdmTMB_sd$b_j, 
    years, sdmTMB_est$b_j + 2 * sdmTMB_sd$b_j,
    col = "red")
  legend("topright", legend = c("VAST", "sdmTMB"), 
    col = c("black", "red"), bty = "n", lty = c(1, 1))
}
```

We wrote some custom code to extract comparable parameters (not shown above). Here are the annual mean estimates in link space with 95% confidence intervals for the two components to the delta model:

```{r compare-vast-sdmTMB2, echo = TRUE}
plot_betas(fit, m1, "beta1_ft")
plot_betas(fit, m2, "beta2_ft")
```

While making custom plots of individual elements would require considerable additional code to extract and reformat the necessary components of each output, VAST has a wrapper function that generates the typical plots one may want. Here we stick with the default set of plots (`plot_set = 3`); however, one can specify different standard plots to make by changing the setting of this argument (see `?FishStatsUtils::plot_maps` and `?FishStatsUtils::plot_results`).

```{r save-plots, echo = TRUE, warning = FALSE, cache=TRUE}
plot(
  fit,
  check_residuals = FALSE,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

@@

```{r plot-knots, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST", "Data_and_knots.png"))
```

@@

```{r plot-density, echo = TRUE, out.width="6in"}
knitr::include_graphics(here("doc", "appendix-VAST", "ln_density-predicted.png"))
```

@@

We can compare the index we would get using sdmTMB. Since sdmTMB does not have built-in delta models, we need some way of combining the predictions across the two models. Here, we will do that by simulating from the joint parameter precision matrix and calculating the index with those draws. This should be approximately comparable with the bias-corrected index calculated above with VAST.

```{r sdmTMB-sim-index, cache=TRUE, echo=TRUE}
set.seed(28239)
p_bin <- predict(m1, newdata = qcs_grid, sims = 1000)
p_pos <- predict(m2, newdata = qcs_grid, sims = 1000)

p_bin_prob <- plogis(p_bin)
p_pos_exp <- exp(p_pos)
p_combined <- log(p_bin_prob * p_pos_exp)
ind_combined_sims <- get_index_sims(p_combined, area = rep(4, nrow(p_combined)))
```

Now, we can compare the indices. We will read in the VAST index from the `Index.csv` file created by the `plot.fit_model()` method above.

```{r index-compare, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
vast_i <- read.csv(here("doc/appendix-VAST/Index.csv")) %>%
  mutate(index = "VAST", year = as.numeric(Time), est = Estimate, 
    se = Std..Error.for.ln.Estimate.) %>% 
  select(index, year, est, se) %>% 
  mutate(lwr = exp(log(est) + qnorm(0.025) * se)) %>% 
  mutate(upr = exp(log(est) + qnorm(0.975) * se))
sdm_i <- ind_combined_sims %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.3) +
  geom_line() + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE)
```
